---
title: "lab_02"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? 

**A1. 32 departments have courses with climate in the title and at least one seat. Atmospheric and Oceanic Science has the most courses with climate in the title.**

```{r}
umd_courses |> 
  mutate(
    title_upper = str_to_upper(title)
  ) |>   
  filter(str_detect(title_upper, "CLIMATE") & seats >= 1) |> 
  group_by(department) |>
  summarise(
    climate_courses = n()
  ) |>
  arrange(desc(climate_courses))
  
  
  # filter(str_detect(title_upper, "CLIMATE") & seats >= 1 & str_detect(department, "German")) |>
  # summarise(
  #   title,
  #   department,
  #   term
  # )

```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why? 

**A2. Germanic studies having a climate class was pretty interesting. I looked into it and it's about "The Great Derangement" which is a fascinating name for anything. The code I used to isolate the Germanic courses also included German Studies courses (string detect with the string being "German") and it's three of the same course. The most recent one is under the German Studies department (Fall 2024) and the other two are under Germanic Studies (Fall 2021 and Fall 2020). I wonder what led the department to changing it**


**Q3** Which departments offered the most courses in the Fall 2024 term? And which departments in that term have at least one syllabus posted? Answer both of those questions and then compare the top five departments - how are they different?

**A3. The top 5 offering departments in Fall 2024 were Business and Management, School of Music, Information Studies, English, and Teaching and Learning, Policy and Leadership. If you filter for courses with a syllabus_count of at least 1, 62 departments are removed entirely and the top 5 changes to Business and Management, Economics, School of Music, Information Studies, and Engineering, Professional Masters. Between the two top 5s there are more humanities focused departments when not factoring syllabi.**

```{r}
umd_courses24_syllabus <- umd_courses |> 
  group_by(department) |> 
  filter(term == "202408", syllabus_count >= 1) |> 
  summarise(
    count_classes = n()
  ) |> 
  arrange(desc(count_classes))
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as a new dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name, and we're not looking in payee_name). Write a sentence describing the pattern of spending over time. 

**A4. The spending pattern generally increased leading up to March of 2022. June of 2022 was a significant peak, likely because of the primary at the start of the following month. The amount goes back down until another peak in October, a month before the general election.**

```{r}
maryland_expenses_dates <- maryland_expenses |> 
  mutate(expenditure_date = mdy(expenditure_date))

maryland_expenses_dates |> 
  mutate(month = floor_date(expenditure_date, "month")
  ) |> 
  filter(committee_name == "Moore  Wes For Maryland") |> 
  group_by(month) |> 
  summarise(
    total_spent = sum(amount)
  ) |> 
  arrange(desc(month))
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for all others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe). 

**A5. In-state spending beat out of state spending by over 26 million dollars.**

```{r}
maryland_expenses_location <- maryland_expenses |> 
  mutate(
    spending_location = case_when(
    str_detect(address, "Maryland") ~ "In-state",
    .default = "Out of state"
  )
  )

maryland_expenses_location |>
  group_by(spending_location) |>
  summarise(
    number = n(),
    total_spent = sum(amount, na.rm = TRUE)
  ) |>
  arrange(desc(total_spent))
```

**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab. 

**A6. My question is are there differences in the amount of courses with diversity in the title between terms. The goal was to see if UMD did a pruning similar to the federal government where they just eradicated anything with diversity in the title. There isn't much of a difference between the most pressing period (since the new administration entered office). But changes likely wouldn't have been seen until Fall of 2025 which isn't included in the dataset. The early 2020s saw the highest course count.**

```{r}
umd_courses |> mutate(
    title_upper = str_to_upper(title)
  ) |> 
  group_by(term) |> 
  filter(str_detect(title_upper, "DIVERSITY") & seats >= 1) |> 
  summarise(
    course_count = n()
  ) |> 
  arrange(desc(course_count))
```
