---
title: "lab_12"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   tidytext and our usual libraries

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
library(tidyverse)
library(tidytext)
library(janitor)
library(lubridate)
library(rvest)
```

## Questions

**Q1.** You've been assigned to report a story about the leading reasons that Maryland attorneys get sanctioned by the state for misconduct. The state [publishes lists of sanctions](https://www.courts.state.md.us/attygrievance/sanctions) that contain a short text description about the situation. Load the CSV file in the data folder containing records from fiscal year 2011 onwards. Make a list of unique words from the text column, then following the example in the pre_lab, remove common "stop words" from that list and create a list of the top 10 words containing the percentage of occurrences each word represents. What's the leading word in that answer and, broadly, what do you think the top 10 words describe?

**A1.**
The leading word is failing but I removed it from my list making client the top word. The top 10 words show that in large part, the reasons for sanctions had to do with ethical violations such as trust and conduct. Also considering the frequency of funds, I think mismanagement of money may be a large factor.
```{r}
sanctions <- read_csv("data/md_attorney_sanctions.csv")

unique_words <- sanctions |>  select(text) |>
  unnest_tokens(word, text)

data("stop_words")

stop_words <- stop_words |>
  add_row(word = "respondent") |> 
  add_row(word = "attorney") |> 
  add_row(word = "failing")

top_ten <- unique_words |>
  anti_join(stop_words) |>
  group_by(word) |>
  tally(sort=TRUE) |> 
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

**Q2.** Let's move beyond single words to phrases. Make a list of the top 10 three-word phrases, called trigrams, based on the example from the pre_lab (you'll need to modify the example code to do this). What's the top trigram and how often does it appear? What does that phrase mean in legal terms?

**A2.**
The top trigram is "conduct involving dishonesty" and it appears 181 times (~3.2%). In legal terms this means the attorney willfully did something deceitful, misleading or generally untruthful.

```{r}
trigrams <- sanctions |> 
  unnest_tokens(trigram, text, token = "ngrams", n = 3) |>
  separate(trigram, c("word1", "word2", "word3"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |>
  filter(!word3 %in% stop_words$word) |>
  mutate(trigram = paste(word1, word2, word3, sep=" ")) |>
  group_by(trigram) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

**Q3.** Let's drop back down to more traditional text analysis - take the top trigram from Q2 and write code to see how many times it occurs in the text column in each fiscal year. What do you think the answer produced by your code suggests? What else could you do to try and clarify the most important reasons attorneys get sanctioned?

**A3.**
I think the results suggest that detection of dishonesty has increased over the past decade and a half, with a massive jump from 2018-2019. After a run peaking in 2021 at 33 instances, the rate has returned slightly above pre-2019 levels. This could suggest an increase in dishonesty or, more charitably, improvements in reporting on dishonesty among attorneys, with the decrease in the past few years showing either the improvements acted as an effective deterrent or reporting has weakened. I think at this point I would do research into any what might've happend in 2019 to cause the spike by speaking to Maryland attorneys that were around then.
```{r}
top_tri_freq <- sanctions |> 
  filter(str_detect(text, "conduct involving dishonesty")) |> 
  mutate(
    frequency_in_text = str_count(text,"conduct involving dishonesty")
  ) |> 
  group_by(fiscal_year) |> 
  summarise(
    frequency = sum(frequency_in_text)
  )
```
