---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```

### Read rds files (binary file type for R)
```{r}
read_rds()
```

### Peak at data as a list of the columns, the data type for each column and and the first few values for each column
```{r}
glimpse()
```

### Peak at the data headers and first 6 rows of data
```{r}
head()
```

### calculate new metrics in a new field using existing fields of data
```{r}
mutate()
```

### get summary statistics of data
```{r}
summary(data)
```


### list columns 
```{r}
colnames(data)
```

### filter data; below would only show courses where the department column is "Journalism"
```{r}
journalism_courses <- umd_courses |> filter(department == "Journalism")
```

### choose what columns should be apart of the object
```{r}
select()
```

### search for pattern in string
```{r}
filter(str_detect(title, "Shakespeare"))
```

### set date column to proper format
```{r}
maryland_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))
#mdy being the key here
```

### string length
```{r}
filter(str_length(`Vendor Zip`) < 5)
```

### standardize column names
```{r}
clean_names(data)
```

### check for dupes
```{r}
data |> get_dupes()
```

### case when
```{r}
maryland_expenses_location <- maryland_expenses |> 
  mutate(
    spending_location = case_when(
    str_detect(address, "Maryland") ~ "In-state",
    .default = "Out of state"
  )
  )
```


### filter by starting with (^)
```{r}
filter(str_detect(data, "^w|^W"))
```

### key collision for making things alike: "The key collision merge function takes each string and extracts the key parts of it. It then puts every key in a bin based on the keys matching."
```{r}
cleaned_md_grant_loans <- md_grant_loans |>
  mutate(grantee_clean=key_collision_merge(Grantee)) |>
  select(Grantee, grantee_clean, everything())

cleaned_md_grant_loans
```

## Another merging algorithim n_gram_merge()
```{r}
cleaned_md_grant_loans <- md_grant_loans |>
  mutate(grantee_clean=n_gram_merge(Grantee)) |>
  select(Grantee, grantee_clean, everything())
```

### Some pdf things
```{r}
fentanyl_deaths <- read_csv("tabula-Annual_2020_Drug_Intox_Report.csv", col_names = FALSE) |> 
  clean_names() |> 
  rename(jurisdiction = x1, deaths_2011 = x2, deaths_2012 = x3, deaths_2013 = x4, deaths_2014 = x5, deaths_2015 = x6, deaths_2016 = x7, deaths_2017 = x8, 
         deaths_2018 = x9, deaths_2019 = x10, deaths_2020 = x11, deaths_total = x12) |> 
  mutate(jurisdiction = str_squish(str_replace_all(jurisdiction,'\\.','')))
```

### bind rows - combines data into a single frame
```{r}
# bind_rows with list
county_voters_combined <- bind_rows(list(county_voters_2016, county_voters_2018, county_voters_2020, county_voters_2022, county_voters_2024))
View(county_voters_combined)
```

# join - combines table by common element
```{r}
maryland_population_with_voters_and_demographics <- maryland_population_with_voters |> left_join(maryland_demographics, join_by(COUNTY==NAME))
```

#change column type
```{r}
maryland_zcta <- maryland_zcta |> mutate(across(ZCTA5N, as.character))
```

#initialize gg plot
```{r}
baltco_911_calls_by_month |>
  ggplot()
```

#basic bar graph (x = x-axis, weight = y-axis)
```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=month, weight=total_calls))
```

#Sort graph from least to greatest
```{r}
x=reorder(month,total_calls)
```

#Flip chart
```{r}
+ coord_flip()
```

#change labels
```{r}
labs(
    title="More 911 Overdose Calls in Warmer Months",
    x = "month",
    y = "total calls",
    caption = "source: Baltimore County"
    
  )
```

#line graph
```{r}
baltco_911_calls_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_calls))
```

#scale date axis
```{r}
 scale_x_date(date_breaks = "1 week", date_labels = "%b %d")
```

#reangle labels
```{r}
theme(
    axis.text.x = element_text(angle = 45,  hjust=1)
  )
```

