---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```

### Read rds files (binary file type for R)
```{r}
read_rds()
```

### Peak at data as a list of the columns, the data type for each column and and the first few values for each column
```{r}
glimpse()
```

### Peak at the data headers and first 6 rows of data
```{r}
head()
```

### calculate new metrics in a new field using existing fields of data
```{r}
mutate()
```

### get summary statistics of data
```{r}
summary(data)
```


### list columns 
```{r}
colnames(data)
```

### filter data; below would only show courses where the department column is "Journalism"
```{r}
journalism_courses <- umd_courses |> filter(department == "Journalism")
```

### choose what columns should be apart of the object
```{r}
select()
```

### search for pattern in string
```{r}
filter(str_detect(title, "Shakespeare"))
```

### set date column to proper format
```{r}
maryland_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))
#mdy being the key here
```

