---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)
# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=369008)

payments <- payments |> 
  clean_names()

# Load in the Maryland grants and loan data]
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")

md_grants_loans <- md_grants_loans |> 
  clean_names() |> 
  rename(source = grantor) |> 
  mutate(source = str_to_upper(source), grantee = str_to_upper(grantee), description = str_to_upper(description))
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: I found 98 records

```{r}
pg_schools <- payments |> 
  filter(str_detect(vendor_name, "PRINCE GEORGE") | str_detect(vendor_name, "PG")) |>  
  filter(str_detect(vendor_name, "EDUCATION") | str_detect(vendor_name, "SCHOOL"))
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: 2024 had the largest total amount at $1.782 billion. It's really weird how there are no payments recorded for 2023. 

```{r}
pg_schools |> 
  group_by(fiscal_year) |> 
  summarise(
    total_payments = sum(amount)
  ) |> 
  arrange(desc(fiscal_year))
```
# Razak tries figuring out if there was a minor mispelling for all the funidng PGCPS received in 2023.
```{r}
payments |> 
  filter(agency_name == "MARYLAND STATE DEPARTMENT OF EDUCATION")
```
#Razak quit trying...



Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3.
1. FISHER SCIENTIFIC @ 02241 and FISHER SCIENTIFIC COMPANY LLC @ 02241
2. DPR CONSTRUCTION @ 20190 and DP CONSTRUCTION @ 21046
3. HOLDER CONSTRUCTION GROUP LLC @ 30339 and HOLDER CONSTRUCTION GROUP, LLC @ 30339

I feel like correcting these would probably have to be manual to some extent. Any algorithm you can make could make a major mistake. A good first step to correct the issue with example #1 would be to show all vendors that have matching zip codes and then you can manually go through and see if any should be merge. For a problem like #2 you can do the same thing but with matching titles. On the off chance that there's a situation where the zip code doesn't match and the name doesn't match in a similar fashion as #3 (one character off), you could try doing a version of string detect that has some sort of percentage slider. So filtering any vendor names that have at least 90% of it's name in common with another vendor.
```{r}
UMD_capital_projects <- payments |> 
  filter(agency_name == "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND") |> 
  group_by(vendor_name, vendor_zip) |> 
  summarise(
    total_received = sum(amount)
  ) |> 
  arrange(desc(total_received))
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: After a massive increase in funding in 2010, STEM Cell research funding peaked in 2012 before falling until 2018. There was a small push into the 2020s that fell in 2021 before rebounding to pre-covid levels in 2022. 2023 became the biggest year in stem cell research funding, passing the $10 million mark.

(2009 is the peak if you add a grant to John Hopkins that also includes other items in its description)

```{r}
md_grants_loans |> 
  filter(str_detect(description, "STEM CELL RESEARCH GRANT")) |> 
  group_by(fiscal_year) |> 
  summarise(
    total = sum(amount)
  )
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5:
There are 47 grantees at the zipcodes associated with the University of Maryland-College Park. To clean the results I would first have to make sure that things I would consider to be the same are actually the same. Intuition tells me University of Marlyand College Park and University of MD College Park are the same thing but I've been wrong before. Also would something like the Clarice fall under the entire school or are they seperate? Then combining and removing "INC" and "LLC" duplicates is a pretty reliable method. Also what is Green Branch Management group and how are they getting more money than the university at this zip code that's pretty insane. Of course the potential duplicates make the university go over Green Branch but still $5.5 million is a lot of money for something I haven't heard of.

```{r}
md_grants_loans |> 
  filter(zip_code == "20740" | zip_code == "20742") |> 
  group_by(grantee) |> 
  summarise(
    total = sum(amount)
  ) |> 
  arrange(desc(total))
```
