---
title: "lab_06"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# load your libraries
library(tidyverse)
library(janitor)
library(lubridate)
```

## Getting Started

This week we'll be combining and joining data to make it more useful and to ask some questions about it. We'll use some more 911 overdose call data to do this. The first thing we want to do is to combine multiple counties' data into a single dataframe so that we can ask some questions. First, let's combine data from Cecil, Carroll and Allegany counties into a new dataframe.

*Before* you combine them you'll need to clean up some of the column names to make the data easier to work with - make sure you have loaded the library to do that. You also need to make sure that each column has the same name and datatype (you can check the datatypes using `glimpse`). If any of them does not, you need to fix that.

```{r}
# load and combine the call data from those three counties
allegany911 <- read_csv("data/allegany_911.csv") |> clean_names()
cecil911 <- read_csv("data/cecil_911.csv") |> clean_names()
carroll911 <- read_csv("data/carroll_911.csv") |> clean_names()

combined911 <- bind_rows(allegany911, carroll911, cecil911)

```

Then we'll join some demographic data with Baltimore City 911 calls and ask some questions of that. For that, you'll need to load the Baltimore City 911 data and the ZCTA demographic data from the data folder, and join them based on the zip code. Remember, the zip codes in both dataframes must be the same datatype (and should be characters). You should start with your Baltimore City 911 dataframe in creating your join.

```{r}
# load and join the Baltimore-related data
baltimoreCity911 <- read_csv("data/baltimore_911.csv")
maryland_zcta <- read_csv("data/maryland_zcta.csv") |> rename(zip_code = ZCTA5N)

glimpse(baltimoreCity911)
glimpse(maryland_zcta)

baltimore911Demographic <- baltimoreCity911 |> left_join(maryland_zcta, by = "zip_code")
```

## Answer questions

Q1. Let's start with our data from Allegany, Cecil and Carroll counties. Write code to generate a dataframe that shows the total number of calls for each county. What's the order of counties from most calls to least?

A1.
The order of counties from most calls to least is Carroll, Cecil, and then Allegany.
```{r}
combined911 |> 
  group_by(county) |> 
  summarise(
    totalCalls = n()
  ) |> 
  arrange(desc(totalCalls))
```

Q2. In your combined dataframe that you used in A1, add a column for the month of each call, then write code to show the total number of calls per county and month. Are there any outliers or noteworthy results? Describe the general pattern of the results.

Then calculate the most calls per county and location. Which location had the most calls - where and what is that location?

A2.
The summer months have the highest overdose calls with August being the highest call month for each county. Allegany's calls are condensed to these months and top the charts, despite having the lowest overall overdose calls. Cecil county had the most total calls in a month at 55 in August. Winter months like December and January have the lowest call rates, with Allegany having the least amount of calls in a month in November. 

314 Grove Neck Rd, Earleville in Cecil County has the most calls at 15, doubling the second highest location. It appears that the locatoin is home to rehab center.
```{r}
combined911 |> 
  mutate(month = month(date, label = TRUE)) |> 
  group_by(county, month) |> 
  summarise(
    totalCalls = n()
  ) |> 
  arrange(desc(totalCalls))

combined911 |> 
  group_by(county, location) |> 
  summarise(
    totalCalls = n()
  ) |> 
  arrange(desc(totalCalls))
```

Q3. Let's turn to the Baltimore City data. You're interested in exploring calls in zip codes with a high percentage of children under 18. Write code to isolate calls in zip codes where the percentage of under 18 population is at least 75% and then show the zip code and population and how many calls occurred in each of those zip codes.

Where are those zip codes in the city?

Add a column to your code that calculates the number of calls per 1,000 people for those zip codes. To calculate a per capita rate, you might find this short guide handy: [https://observablehq.com/\@palewire/per-capita-calculator](https://observablehq.com/@palewire/per-capita-calculator){.uri}.

Which zip code has the highest rate of calls per 1,000 people? Find the neighborhoods in that zip code that are listed in the data - you can use group_by or distinct to do this. What are some of those neighborhoods, and what else can you tell me about the population there?

A3.
21217, 21223, 21202, 21213, 21201, 21216, 21205

21205 has the highest rate of calls per 1,000 people. The top 3 neighborhoods by call volume are McElderry Park, Milton-Montford and Gay Street.



```{r}
baltimore911Demographic |> 
  filter(PCT_UNDER_18 > 75) |> 
  group_by(zip_code) |> 
  summarise(
    population = sum(POPULATION),
    totalCalls = n(),
    perCapita = (totalCalls/population) * 1000
  ) |> 
  arrange(desc(perCapita))

baltimore911Demographic |> 
  filter(zip_code == 21205) |> 
  group_by(neighborhood) |> 
  summarise(
    totalCalls = n()
  ) |> 
  arrange(desc(totalCalls))
```

Q4. Choose your own adventure: I want you to come up with a question you could ask of the Baltimore City data that combines both the calls and the demographic information. You must use both demographics and the overdose call data in your work.

A4:
A question I could ask the Baltimore City data is how many overdoses occurred in areas that are at least 50% white. Based on the results 83% of all overdose calls came from zip codes with over half the population being non-white.
```{r}
baltimore911Demographic |> 
  group_by(PCT_WHITE_NON_HISPANIC >= 50) |> 
  summarise(
    total = n(),
    perct = total/nrow(baltimore911Demographic)
  )
```
