---
title: "lab_01"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you? **A1**

The department with the highest mean is Computer, Math, and Natural Sciences. It's not super helpful because it's for a class that's only been offered seemingly once. I think it's intersting how large Air Science classes are. It's not my first thought when it comes to popular subjects but it's in the top 10.

```{r}
umd_courses |> 
  group_by(department) |> 
  summarise(
    totalSeats = sum(seats),
    averageSeats = mean(seats)
  ) |> 
  arrange(desc(averageSeats))
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your code must show the greatest number of classes first.
**A2**
The greatest number of times a class has been offered is 29 times.
```{r}
umd_courses |> 
  group_by(id) |> 
  summarise(
    timesOffered = n()
  ) |> 
  arrange(desc(timesOffered))
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most?
**A3**
Jason Kuo has taught the most courses as the only section offering the class. NA and TBA classes are also popular.

```{r}
umd_courses |> 
  group_by(instructors) |> 
  summarise(
    timesTaught = n()
  ) |> 
  arrange(desc(timesTaught))
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? **A4**
Having the data be separated by section offerings rather than the aggregate by courses each term would have made it possible to see who has really taught the most classes. The current data puts all the instructors into one cell so if Jason Kuo taught a class along with Manoj Franklin, it wouldn't be represented in the 94 gotten in the previous answer. 


**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? **A5**
My question was: 
What department has the smallest instructor to available seats ratio?

I can't breakdown the indvidual instructors (so if CMNS100 is taught by 50 people, they're all count as one) so instead I found the course to student ratio. Less useful but still kind of interesting. Originally I only grouped by department and got the CMNS class from earlier that dominated the dataset because it was offered once and had so many students. Grouping by term showed that summer and winter semesters have the highest ratios. 


```{r}
umd_courses |> 
  group_by(department) |> 
  summarise(
    totalSeats = sum(seats),
    totalCourses = n(),
    seatsPerCourse = totalSeats/n()
  ) |> 
  arrange(desc(seatsPerCourse))
```
